---
title: "PI Curve Raw Data Processing"
author: "Micaela Chapuis"
date: "2026-09-04"
output: html_document
---
## Respo Code for Processing Raw PI Curve Data
### Created by: Nyssa Silbiger, updated by Maya Powell

This code is run once per PI Curve run to create a file with respo outputs. 

## Install Packages (if not already installed)
```{r message = FALSE}
if ("LoLinR" %in% rownames(installed.packages()) == 'FALSE') devtools::install_github('colin-olito/LoLinR')
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse')
if ("here" %in% rownames(installed.packages()) == 'FALSE') install.packages('here')
if ("future" %in% rownames(installed.packages()) == 'FALSE') install.packages('future')
if ("furrr" %in% rownames(installed.packages()) == 'FALSE') install.packages('furrr')
```

## Load Libraries
```{r message = FALSE}
library(tidyverse) 
library(here) 
library(LoLinR)
library(future) 
library(furrr) 
```

### Set file path 
Set file path to raw oxygen datasheets
```{r}
path_o2 <- here("Data","Respo","Pilot", "PI Curves") # the location of all respirometry files
```

Bring in all of the individual files

```{r}
filenames <- basename(list.files(path = path_o2, pattern = "csv$", recursive = FALSE)) # list all csv file names in the folder 
# recursive = FALSE means it will not list file in subfolders, change to TRUE if that is needed
```

If going into subfolders:
Basename above removes the subdirectory name from the file, storing the full name here
```{r}
#filenames_full <- list.files(path = path_o2, pattern = "csv$", recursive = TRUE) 
```

### Load in Data
Loading in respo metadata (file with times, light levels, etc)
```{r message = FALSE}
sample_data <-read_csv(here("Data", "Respo", "Pilot", "pilot_PI_curves.csv"))
```


## Make sure times are consistent
Make start and stop times real times, so that we can join the respo output and sample_data data frames
```{r}
sample_data <- sample_data %>% 
  unite(respo_date,start_time,col="start_time",remove=F, sep=" ") %>% # makes datetime column for respo start
  unite(respo_date,end_time,col="end_time",remove=F, sep=" ") %>% # makes datetime column for respo end
  mutate(start_time = ymd_hms(start_time),
         end_time = ymd_hms(end_time),
         respo_date = ymd(respo_date))
```

## Generate an empty dataframe
Data is in umol/L/sec

```{r}
n_light_levels <- 10 # number of unique light levels

respo_output <- tibble(.rows =length(filenames)*n_light_levels,
                        sample_ID = NA,
                        intercept = NA,
                        umol_liter_sec = NA,
                        temp_C = NA,
                        light_level = NA,
                        average_PAR = NA)
```

### For Loop
```{r}
for(i in 1:length(filenames)) {
  find_row <- as.numeric(which(sample_data$measurement_name == filenames[i])) # Find the row in sample_data where the measurement_name (file name) matches each name from filenames list
 
  respo_data1 <- read_csv(skip=1,file.path(path_o2, paste0(filenames[i]))) %>% # reads in each file in list
                 select(Date, Time, Value, Temp) %>% # keep only what we need from the oxygen files: Time stamp, Raw O2 value, in situ temp
                 unite(Date, Time, col="Time", remove=T, sep = " ") %>% # make datetime column
                 drop_na() %>% 
                 mutate(Time = mdy_hms(Time)) # convert time

# Now cut the data by start and stop times from respo metadata
# Use start time of each light step from the metadata to separate data by light stop
  
  oxy_subsets <- sample_data[find_row,] %>%
      pmap(function(light_level, start_time, end_time, ...) {
      data <- respo_data1  %>%
              filter(Time >= start_time & Time <= end_time) %>%
              arrange(Time) %>%
              mutate(t_sec = as.numeric(difftime(Time, first(Time), units = "secs"))) %>% #keep everything in seconds
              mutate(light_level = light_level) %>%
              filter(t_sec > 180) %>%              # drop first 3 min (180 s)
              filter(row_number() %% 5 == 0)      # keep every 5th row (if we need to thin the data)

    }) 
  
  
# Combine all oxy subsets (data subsets per light level) into one long dataframe with ID labels
  combined_oxy <- bind_rows(oxy_subsets)
     
# Get the filename without the _O2.csv
  rename <- sub("_O2.csv","", filenames[i])
  
# Plot and export the thinned data
  p1 <- ggplot(combined_oxy, aes(x = t_sec, y = Value)) +
    geom_point(color = "dodgerblue") +
    labs(
      x = 'Time (seconds)',
      y = expression(paste(' O'[2],' (',mu,'mol/L)')),
      title = "original"
    ) +
    facet_wrap(~light_level, scales = "free_y")
  
  
##Olito et al. 2017: It is running a bootstrapping technique and calculating the rate based on density
#option to add multiple outputs method= c("z", "e "pc")
  
# LoLinR = Local Linear Regression in R --> will run multiple linear regressions and suggest the best possible slope
  
# Define function for fitting LoLinR regressions to be applied to all intervals for all samples
  fit_reg <- function(data) {
    rankLocReg(xall = data$t_sec, yall = data$Value, 
               alpha = 0.2, method = "pc", verbose = FALSE)
  }
  
   
# Setup for parallel processing
  future::plan(multisession)
  
# Map LoLinR function onto all intervals of each sample's thinned dataset
  df <- combined_oxy %>%
    select(t_sec, Value, light_level, Temp) %>%
    mutate(t_sec = as.numeric(t_sec)) %>%
    nest_by(light_level) %>%
    ungroup() %>%
    mutate(regs = furrr::future_map(data, fit_reg), # run the LoLinR fit in parallel
           temp_C = map_dbl(map(data, "Temp"), mean), # get the mean temperature
           RegStats = map(regs, function(x){ # extract the intercept and slope for the parameters
                          x$allRegs %>%
                          slice(1) %>%
                          select(intercept = b0,
                                 umol_liter_sec = b1)
           }) )

    
# Plot regression diagnostics and save them in Output folder
 
  for(j in 1:length(df$light_level)){
   pdf(paste0(here("Output","Pilot", "PI Curves"),"/",rename,"_",j,".pdf" ))
    plot(df$regs[[j]])
    dev.off() 
  }
 
  
  df <- df %>%
    select(light_level, temp_C, RegStats) %>%
    unnest(RegStats) %>%
    mutate(sample_ID = rename) %>%
    left_join(sample_data[find_row,] %>%
              select(sample_ID, light_level, average_PAR)) # make sure the important metadata is in the final DF
    

# Fiill in all the O2 consumption and rate data
  
  respo_output[find_row,"temp_C"] <- df$temp_C
  respo_output[find_row,"sample_ID"] <- df$sample_ID
  respo_output[find_row,"intercept"] <- df$intercept
  respo_output[find_row,"umol_liter_sec"] <- df$umol_liter_sec
  respo_output[find_row,"light_level"] <- df$light_level
  respo_output[find_row,"average_PAR"] <- df$average_PAR

 }  
```

### Export Data
```{r}
write_csv(respo_output, here("Data", "Respo", "Pilot", "respo_PI_output.csv"))
```

