---
title: "PI Curve Analysis"
author: "Micaela Chapuis"
date: "2026-09-04"
output: github_document
---

# Respo Code for PI Curve Analysis
### Created by: Nyssa Silbiger, updated by Maya Powell

This code takes the output from the PI_curve_processing code, as well as information about the PI run and tile/chamber measurements, and: 
 
- calculates rates  
- blank-corrects them  
- normalizes them to surface area and dry weight  
- plots PI curve   
- calculates Ik   

## Load Libraries

```{r message = FALSE}
library(segmented)
library(plotrix)
library(gridExtra)
library(LoLinR)
library(lubridate)
library(chron)
library(patchwork)
library(tidyverse)
library(here)
library(PNWColors)
library(ggrepel)
library(reshape2)
library(viridis)
library(car)
library(future)
library(furrr)
library(nls.multstart)
library(ggpubr)
library(investr)
```

## Load Data

```{r message = FALSE}
respo_data <-read_csv(here("Data", "Respo", "Pilot", "pilot_PI_curves.csv"))
tile_measurements <- read_csv(here("Data","Respo","tile_measurements.csv"))
respo_output <- read_csv(here("Data", "Respo", "Pilot", "respo_PI_output.csv"))
```

## Join respo metadata and tile measurements

```{r}
sample_data <- left_join(respo_data, tile_measurements)
```

### Calculate Respiration Rate

```{r}
respo_rates <- respo_output %>%
  left_join(sample_data) %>% # Join the raw respo calculations with the metadata
  mutate(chamber_volume_l = chamber_volume_ml * 0.001,  # mL to L conversion
         umol_sec = umol_liter_sec * chamber_volume_l) %>% # Account for chamber volume to convert from umol L-1 s-1 to umol s-1. This standardizes across water volumes (different because different communities on each tile, thickness of epoxy, etc) and removes per Liter
  mutate_if(sapply(., is.character), as.factor)  # convert character columns to factors
  
#Account for blank rate by sample run Block (if we do at least one blank per block)

```

### Normalize Respiration Rates to Blanks

```{r message = FALSE}
rates_normalized <- respo_rates %>% 
  filter(blank == 1) %>% # grab the blanks
  group_by(light_level, run_block, light_dark) %>%
  summarise(blank_rate = mean(umol_sec, na.rm = TRUE)) %>% # if you have multiple blanks per run take the average
  ungroup() %>% 
  #dplyr::select(Light_level, run_block, blank_rate, date) %>% # this is what I will use to join the blanks back with the raw data
  right_join(respo_rates) %>% # join blanks with the respo data
  mutate(
      # Blank corrected rates
        umol_sec_corrected = umol_sec - blank_rate, # Blank corrected rates = subtract the blank rates from the raw rates   
      # Rates in micromols, normalized to surface area (in cm2)   
         umol_cm2_hr_uncorrected = (umol_sec*3600)/surface_area_cm2,
         umol_cm2_hr = (umol_sec_corrected*3600)/surface_area_cm2, 
      # Rates in micromols, normalized to dry weight (in g)
         umol_g_hr_uncorrected = (umol_sec*3600)/dry_weight_g, 
         umol_g_hr = (umol_sec_corrected*3600)/dry_weight_g,
      # Rates in millimols, normalized to surface area (in cm2)
         mmol_cm2_hr_uncorr = 0.001*(umol_sec*3600)/surface_area_cm2,
         mmol_cm2_hr = 0.001*(umol_sec_corrected*3600)/surface_area_cm2, # convert to mmol cm-2 hr-1, normalized to surface area
      # Rates in millimols, normalized to dry weight (in g)
         mmol_g_hr_uncorr = 0.001*(umol_sec*3600)/dry_weight_g,
         mmol_g_hr = 0.001*(umol_sec_corrected*3600)/dry_weight_g
        ) %>% 
  filter(blank!=1) %>% # remove the Blank data
  select(respo_date, sample_ID, tile_ID, treatment, light_dark, run_block, surface_area_cm2, dry_weight_g, run_block, umol_cm2_hr,umol_cm2_hr_uncorrected, umol_g_hr, umol_g_hr_uncorrected, mmol_cm2_hr_uncorr, mmol_cm2_hr, mmol_g_hr_uncorr, mmol_g_hr, channel_number, temp_C, light_level, average_PAR) #keep only what we need


```

## Export df

```{r}
write_csv(rates_normalized, here("Data","Respo", "Pilot", "PI_rates_normalized.csv"))  
```

## Making a dataframe for blank data for future use in plots

```{r message = FALSE}
blanks_only <- respo_rates %>% 
  filter(blank == 1) %>% # grab the blanks
  group_by(light_level, average_PAR, run_block, light_dark) %>%
  summarise(blank_rate = mean(umol_sec, na.rm = TRUE))
```

## Plot the blanks across treatments to make sure everything looks good

```{r}
blanks_only %>%
  ggplot(aes(x = average_PAR, blank_rate)) +
  geom_point()

# if i had a big outlier, then I would take avg of the blanks and use that for the blanks
# my blanks look like that because I have a lot of water, but the range/spread on the Y axis is very small (0.003)
```

## Basic plot of rates versus light before making real PI curve

Normalized to surface area

```{r}
rates_normalized %>%
  ggplot(aes(x = average_PAR, y = umol_cm2_hr, color = tile_ID))+
  geom_point() +
  geom_line()
#ggsave(here("Output","PI","basic_PI_plot.pdf"), basic_PI_plot)

# curves should always start below 0 because we start in the dark
# when curves start going down, it may be photoinhibition

# overall P rates similar across tiles
```

Normalized to dry weights

```{r}
rates_normalized %>%
  ggplot(aes(x = average_PAR, y = umol_g_hr, color = tile_ID))+
  geom_point() +
  geom_line()
#ggsave(here("Output","PI","basic_PI_plot.pdf"), basic_PI_plot)

# normalized rates v different between tiles, opposite pattern than i would expect. Would expect turfy highest? 
# weight : photosynthesis ratio for encrusting is higher?
# so maybe green is photoinhibition at 500
# stick to under 500, maybe 450
```

### Run NLS model for PI curve and extract Ik for each tile

##### Nonlinear Least Squares regression of a non-rectangular hyperbola (Marshall & Biscoe, 1980)

```{r}
# set treatment colors
treatment_colors <- c("black" = "#698B69",
                      "white" = "#96CDCD",
                      "white_copper" = "#CD5B45")
```

```{r}
#define nrh model & function to make sure all variables are not saved from previous runs
nrh_model <- function(average_PAR, Am, AQY, Rd, theta) {
  (1/(2*theta)) * (AQY*average_PAR + Am - sqrt((AQY*average_PAR + Am)^2 - 4*AQY*theta*Am*average_PAR)) - Rd
}

make_nrh_function <- function(Am, AQY, Rd, theta) {
  force(Am); force(AQY); force(Rd); force(theta)
  function(x) nrh_model(x, Am, AQY, Rd, theta)
}
```

```{r}
#create plot list and a coefficient table list
plot_list <- list()
coef_list <- list()
```

```{r}
#for loop to calculate ik and plot
#Set i =  to whatever you want to look at for each treatment as you go through
for (i in unique(rates_normalized$treatment)) {
  data <- rates_normalized %>% filter(treatment == i)
  
  # use nls_multstart to calculate the curve
  curve_nlslrc <- nls_multstart(
    # using corrected rates
    umol_cm2_hr ~ nrh_model(average_PAR, Am, AQY, Rd, theta), 
    data = data,
    iter = 500,
    
    #Am (max photosynthetic rate) should be positive based on range of umol_cm2_hr - here ~0-2
    #AQY (apparent quantum yield) should be positive and between 0-1 (often much smaller)
    #Rd (dark respiration) should be negative and usually between -2-0 but can adjust if larger
    #theta (curvature of the line) should be positive, between 0 and 1 
    
    #IF YOU ARE NORMALIZING TO G INSTEAD OF CM2 OR ANOTHER METRIC YOU SHOULD LOOK AT YOUR DATA FIRST TO SEE BOUNDS FOR AM, AQY AND RD
    
    start_lower = c(Am = 0,     AQY = 0,    Rd = -2,  theta = 0),
    start_upper = c(Am = max(data$umol_cm2_hr) * 2, AQY = 1,    Rd = 0,    theta = 1),
    lower = c(Am = 0, AQY = 0, Rd = -Inf, theta = 0),
    supp_errors = 'Y'
  )
  
  # create 95% CI
  light_seq <- seq(min(data$average_PAR), max(data$average_PAR), length.out = 200)
  
  ci_pred <- predFit(curve_nlslrc, 
                     newdata = data.frame(average_PAR = light_seq), 
                     interval = "confidence", 
                     level = 0.95)
  
  ci_df <- data.frame(average_PAR = light_seq,
                      fit = ci_pred[, "fit"],
                      lwr = ci_pred[, "lwr"],
                      upr = ci_pred[, "upr"])
  
  coefs <- coef(curve_nlslrc)
  
  coef_print <- coefs %>%
    t() %>%
    as.data.frame() %>%
    rename(Pmax_gross = Am) %>% # max photosynthetic rate
    mutate(Ik = Pmax_gross/AQY, # calculate Ik
           Ic = Rd/AQY,
           Pmax_net = Pmax_gross - Rd,
           treatment = i)
  
  # get Ik value (and force so that it updates each loop)
  ik_val <- coef_print$Ik
  force(ik_val)
  
  fitted_function <- make_nrh_function(coefs["Am"], coefs["AQY"], coefs["Rd"], coefs["theta"])
  
  
  # make plot
   basic_PI_plot <- data %>%
    ggplot(aes(x = average_PAR, y = umol_cm2_hr, color = treatment_colors[i]))+
    geom_point(shape = 21)+
    #geom_line()+
    geom_ribbon(data = ci_df, aes(x = average_PAR, ymin = lwr, ymax = upr), 
                inherit.aes = FALSE, fill = treatment_colors[i], alpha = 0.2) +
    theme_classic(base_size = 15) +
    geom_vline(xintercept = ik_val, color = treatment_colors[i], lty = 2, linewidth = 1) +
    annotate("text", x = ik_val + 100, y = 0, label = paste0("Ik = ", round(ik_val))) +
    #you can change ik plus however much you need to get the spacing correct
    stat_function(fun = fitted_function, linewidth = 1, color = treatment_colors[i]) +
    #add in 95% CI to line
    labs(x = expression("Photon Flux Density ("*mu*"mol photons "*m^-2*s^-1*")"),
         y = expression(Rate*" ("*mu*"mol "*O[2]*" "*cm^-2*h^-1*")"),
         title = i)+
    theme(legend.position = "none", plot.title = element_text(face = "italic"))
   
  plot_list[[i]] <- basic_PI_plot
  coef_list[[i]] <- coef_print
  nls_model_coef_all <- bind_rows(coef_list)
  
}
```

```{r}
ggarrange(plotlist = plot_list) #, ncol = 5, nrow = 2)
#ggsave(here("Output/PI/ikonik_publishable_plot.pdf"), all_nls_PI_plots, width = 20, height = 8)
```

```{r}
print(nls_model_coef_all)
```

```{r}
#write.csv(nls_model_coef_all, here("Output/PI/nls_model_coef_all.csv"))
```

### Old/Outdated Code

## Plot curves

```{r}
# aggregating all tiles
## tile_respo <- rates_normalized %>% filter(     ) # in future might want to filter by treatment (black/white)

#tile_mean <- aggregate(umol_g_hr ~ average_PAR, data = rates_normalized, FUN=mean)
#tile_se <- aggregate(umol_g_hr ~ average_PAR, data = rates_normalized, FUN=std.error)
```

Tile Data

```{r}
#PAR <- as.numeric(rates_normalized$average_PAR)
#rates <- as.numeric(rates_normalized$umol_cm2_hr)
#plot <- plot(PAR, rates,
#     xlab="", ylab="", 
#     xlim=c(0,max(PAR)), #ylim=c(-0.5, 1.2), 
#     cex.lab=0.8, cex.axis=0.8, cex=1, 
#     main="All tiles", adj = 0.05) #set plot info
#mtext(expression("Irradiance ("*mu*"mol photons "*m^-2*s^-1*")"), side=1, line=3.3, cex=1) #add labels
#mtext(expression(Rate*" ("*mu*"mol "*O[2]*" "*cm^-2*h^-1*")"), side=2, line=2, cex=1) #add labels
```

### Fit a model using a Nonlinear Least Squares regression of a non-rectangular hyperbola (Marshall & Biscoe, 1980)

```{r}
#curve_nlslrc = nls(rates ~ (1/(2*theta)) * (AQY*PAR+Am-sqrt((AQY*PAR+Am)^2-4*AQY*theta*Am*PAR)) - Rd,
#                   start = list(Am = (max(rates) - min(rates)),
#                                AQY = 0.15, # might need to change for g
#                                Rd = -min(rates),
#                                theta = 0.2)) # might need to change for g, how quickly it goes up when it starts photosynthesizing (E?)
#fit <- summary(curve_nlslrc) #summary of model fit
#fit
```

```{r}
# plot needs to be in the same chunk to be able to add the curve to it
#plot(PAR, rates,
#     xlab="", ylab="", 
#     xlim=c(0,max(PAR)), #ylim=c(-0.5, 1.2), 
#     cex.lab=0.8, cex.axis=0.8, cex=1, 
#     main="All tiles", adj = 0.05) #set plot info
#mtext(expression("Irradiance ("*mu*"mol photons "*m^-2*s^-1*")"), side=1, line=3.3, cex=1) #add labels
#mtext(expression(Rate*" ("*mu*"mol "*O[2]*" "*cm^-2*h^-1*")"), side=2, line=2, cex=1) #add labels

#tile_curve_fitting <- curve((1/(2*fit$coef[4,1]))*(fit$coef[2,1]*x + fit$coef[1,1] - sqrt((fit$coef[2,1]*x + fit$coef[1,1])^2 - 4*fit$coef[2,1]*fit$coef[4,1]*fit$coef[1,1]*x)) - fit$coef[3,1], 
#                      lwd=2, col="blue", add=T)
```

### Extract values

```{r}
#Amax (max gross photosytnthetic rate)
#Pmax.gross <- fit$parameters[1]

#AQY (apparent quantum yield) alpha
#AQY <- fit$parameters[2]

#Rd (dark respiration)
#Rd <- fit$parameters[3]

# Ik light saturation point
#Ik <- Pmax.gross/AQY

# Ic light compensation point
#Ic <- Rd/AQY

# Net photosynthetic rates
#Pmax.net <- Pmax.gross - Rd
```

## Output parameters into a table

```{r}
#pilot_PI_output <- rbind(Pmax.gross, Pmax.net, -Rd, AQY, Ik, Ic)
#row.names(pilot_PI_output) <- c("Pg.max","Pn.max","Rdark","alpha", "Ik", "Ic")

#pilot_PI_output
```
