---
title: "Respo Data Analysis"
author: "Micaela Chapuis"
date: "2026-10-02"
output: github_document
---

# Code for Light/Dark Respo Analysis
### Created by: Nyssa Silbiger, updated by Maya Powell

This code takes the output from the respo_raw_processing code, as well as information about the respo runs and tile/chamber measurements, and: 

- blank-corrects rates 
- normalizes them to surface area and dry weight 
- calculates respiration, net photosynthesis and gross photosynthesis rates  

  
## Load Libraries
```{r message = FALSE}
library(tidyverse) 
library(here) 
library(patchwork)
```

## Load Data  
```{r message = FALSE}
respo_data <-read_csv(here("Data", "Respo", "Pilot", "pilot_tile_respirometry.csv"))
tile_measurements <- read_csv(here("Data","Respo","tile_measurements.csv"))
respo_output <- read_csv(here("Data", "Respo", "Pilot", "respo_lightdark_output.csv"))
```

## Join respo metadata and tile measurements  
```{r}
sample_data <- left_join(respo_data, tile_measurements, by = join_by(sample_ID, tile_ID))
```

### Calculate Respiration Rate  
```{r}
respo_rates <- respo_output %>%
  left_join(sample_data) %>% # Join the raw respo calculations with the metadata
  mutate(chamber_volume_l = chamber_volume_ml * 0.001,  # mL to L conversion
         umol_sec = umol_liter_sec * chamber_volume_l) %>% # Account for chamber volume to convert from umol L-1 s-1 to umol s-1. This standardizes across water volumes (different because different communities on each tile, thickness of epoxy, etc) and removes per Liter
  mutate_if(sapply(., is.character), as.factor)  # convert character columns to factors
  
#Account for blank rate by sample run Block (if we do at least one blank per block)

```


### Normalize Respiration Rates to Blanks  
```{r message = FALSE}
rates_normalized <- respo_rates %>% 
  filter(blank == 1) %>% # grab the blanks
  group_by(run_block, light_dark) %>% # can't group by light level because they're slightly different for different respo stands
  summarise(blank_rate = mean(umol_sec, na.rm = TRUE)) %>% # if you have multiple blanks per run take the average
  ungroup() %>% 
  #dplyr::select(Light_level, run_block, blank_rate, date) %>% # this is what I will use to join the blanks back with the raw data
  right_join(respo_rates) %>% # join blanks with the respo data
  mutate(
      # Blank corrected rates
        umol_sec_corrected = umol_sec - blank_rate, # Blank corrected rates = subtract the blank rates from the raw rates   
      # Rates in micromols, normalized to surface area (in cm2)   
         umol_cm2_hr_uncorrected = (umol_sec*3600)/surface_area_cm2,
         umol_cm2_hr = (umol_sec_corrected*3600)/surface_area_cm2, 
      # Rates in micromols, normalized to dry weight (in g)
         umol_g_hr_uncorrected = (umol_sec*3600)/dry_weight_g, 
         umol_g_hr = (umol_sec_corrected*3600)/dry_weight_g,
      # Rates in millimols, normalized to surface area (in cm2)
         mmol_cm2_hr_uncorrected = 0.001*(umol_sec*3600)/surface_area_cm2,
         mmol_cm2_hr = 0.001*(umol_sec_corrected*3600)/surface_area_cm2, # convert to mmol cm-2 hr-1, normalized to surface area
      # Rates in millimols, normalized to dry weight (in g)
         mmol_g_hr_uncorrected = 0.001*(umol_sec*3600)/dry_weight_g,
         mmol_g_hr = 0.001*(umol_sec_corrected*3600)/dry_weight_g
        ) %>% 
  filter(blank!=1) %>% # remove the Blank data
  select(respo_date, sample_ID, tile_ID, treatment, light_dark, run_block, surface_area_cm2, dry_weight_g, umol_cm2_hr,umol_cm2_hr_uncorrected, umol_g_hr, umol_g_hr_uncorrected, mmol_cm2_hr_uncorrected, mmol_cm2_hr, mmol_g_hr_uncorrected, mmol_g_hr, channel_number, temp_C, light_level) #keep only what we need
```

## Making a dataframe for blank data for future use in plots  
```{r message = FALSE}
blanks_only <- respo_rates %>% 
  filter(blank == 1) %>% # grab the blanks
  group_by(light_dark, run_block) %>%
  summarise(blank_rate = mean(umol_sec, na.rm = TRUE))
```

## Plot the blanks across treatments to make sure everything looks good  
```{r}
blanks_only %>%
  ggplot(aes(x = light_dark, y = blank_rate)) +
  geom_boxplot(alpha = 0.4) + geom_point()
```


### Respiration, Net + Gross Photosynthesis Rates, Normalized to Surface Area  
```{r}
rates_PR_SAnorm <- rates_normalized %>%
  select(-c(light_level,temp_C, # remove the things that make light and dark runs different and keep them in separate rows
            umol_cm2_hr_uncorrected, umol_g_hr, umol_g_hr_uncorrected, mmol_cm2_hr_uncorrected, mmol_cm2_hr, mmol_g_hr_uncorrected, mmol_g_hr)) %>% # remove uncorrected column, dry weight column, and mmol colums to pivot
  pivot_wider(names_from = light_dark, 
              values_from = umol_cm2_hr) %>%  # makes light and dark columns with the rates for their respective run 
  rename(respiration = dark , net_photosynthesis = light) %>% # rename the columns
  mutate(respiration = -1 * respiration,  # Make respiration positive
         gross_photosynthesis = respiration + net_photosynthesis) %>% 
  pivot_longer(cols = respiration:gross_photosynthesis, names_to = "PR", values_to = "values") #values still in umol.cm2.hr

rates_PR_SAnorm
```

```{r}
write_csv(rates_PR_SAnorm,here("Data","Respo", "Pilot", "Respo Outputs", "PnR_rates_SAnorm.csv")) # export all the uptake rates
```

### Respiration, Net + Gross Photosynthesis Rates, Normalized to Dry Weights  
```{r}
rates_PR_DWnorm <- rates_normalized %>%
  select(-c(light_level,temp_C, # remove the things that make light and dark runs different and keep them in separate rows
            umol_g_hr_uncorrected, umol_cm2_hr, umol_cm2_hr_uncorrected, mmol_cm2_hr_uncorrected, mmol_cm2_hr, mmol_g_hr_uncorrected, mmol_g_hr)) %>% # remove uncorrected column, dry weight column, and mmol colums to pivot
  pivot_wider(names_from = light_dark, 
              values_from = umol_g_hr) %>%  # makes light and dark columns with the rates for their respective run 
  rename(respiration = dark , net_photosynthesis = light) %>% # rename the columns
  mutate(respiration = -1 * respiration,  # Make respiration positive
         gross_photosynthesis = respiration + net_photosynthesis) %>% 
  pivot_longer(cols = respiration:gross_photosynthesis, names_to = "PR", values_to = "values") #values still in umol.cm2.hr

rates_PR_DWnorm
```

```{r}
write_csv(rates_PR_SAnorm,here("Data","Respo", "Pilot", "Respo Outputs", "PnR_rates_DWnorm.csv")) # export all the uptake rates
```

## Plots  
#### Algae Biomass by Treatment  
```{r}
tile_measurements %>%
  filter(treatment %in% c("white", "black")) %>%
  ggplot(aes(x = treatment, y = dry_weight_g, color = treatment)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.4) +
  theme_bw() +
  labs(y = "Dry Algae Biomass (g)") +
  guides(color = "none")
```

```{r}
tile_measurements %>% 
  filter(treatment %in% c("white", "black")) %>% 
  group_by(treatment) %>% 
  summarize(mean_weight = mean(dry_weight_g))
```


#### Photosynthesis + Respiration Plot, Surface Area Normalized  
```{r}
rates_PR_SAnorm %>% 
  ggplot(aes(x = treatment, y = values, color = treatment)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.4) +
  facet_wrap(~PR ,scales = "free") + #*run_block, nrow = 3
  theme_bw() +
  labs(x = "Treatment", y = "umol.cm2.hr")+
  theme(strip.background = element_rect(fill = "white"),
        strip.text = element_text(face = "bold")) +
  guides(color = "none")
```

#### Photosynthesis + Respiration Plot, Dry Weight Normalized  
```{r}
rates_PR_DWnorm %>% 
  ggplot(aes(x = treatment, y = values, color = treatment)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.4) +
  facet_wrap(~PR, scales = "free") +
  theme_bw() +
  labs(x = "Treatment", y = "umol.g.hr")+
  theme(strip.background = element_rect(fill = "white"),
        strip.text = element_text(face = "bold")) +
  guides(color = "none")
```

```{r}
rates_PR_SAnorm %>% 
  group_by(treatment, PR) %>% 
  summarize(mean_value = mean(values))
```

```{r}
rates_PR_DWnorm %>% 
  group_by(treatment, PR) %>% 
  summarize(mean_value = mean(values))
```

#### Gross Photosynthesis Plot, Surface Area Normalized  
```{r}
rates_PR_SAnorm %>% filter(PR == "gross_photosynthesis") %>%
  ggplot(aes(x = treatment, y = values, color = treatment)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.4) +
  theme_bw() +
  labs(y ="Gross Photosynthesis (umol.cm2.hr)") +
  guides(color = "none")
```

#### Gross Photosynthesis Plot, Dry Weight Normalized  
```{r}
rates_PR_DWnorm %>% filter(PR == "gross_photosynthesis") %>%
  ggplot(aes(x = treatment, y = values, color = treatment)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.4) +
  theme_bw() +
  labs(y ="Gross Photosynthesis (umol.g.hr)") +
  guides(color = "none")
```

#### Net Photosynthesis Plot, Surface Area Normalized  
```{r}
rates_PR_SAnorm %>% filter(PR == "net_photosynthesis") %>%
  ggplot(aes(x = treatment, y = values, color = treatment)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.4) +
  theme_bw() +
  labs(y ="Net Photosynthesis (umol.cm2.hr)") +
  guides(color = "none")
```

#### Net Photosynthesis Plot, Dry Weight Normalized  
```{r}
rates_PR_DWnorm %>% filter(PR == "net_photosynthesis") %>%
  ggplot(aes(x = treatment, y = values, color = treatment)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.4) +
  theme_bw() +
  labs(y ="Net Photosynthesis (umol.g.hr)") +
  guides(color = "none")
```

#### Respiration Plot, Surface Area Normalized  
```{r}
rates_PR_SAnorm %>% filter(PR == "respiration") %>%
  ggplot(aes(x = treatment, y = values, color = treatment)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.4) +
  theme_bw() +
  labs(y ="Respiration (umol.cm2.hr)") +
  guides(color = "none")
```

#### Respiration Plot, Dry Weight Normalized  
```{r}
rates_PR_DWnorm %>% filter(PR == "respiration") %>%
  ggplot(aes(x = treatment, y = values, color = treatment)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.4) +
  theme_bw() +
  labs(y ="Respiration (umol.g.hr)") +
  guides(color = "none")
```

#### Rates by Tile
```{r}
rates_normalized %>%
  ggplot(aes(x = tile_ID, y = umol_cm2_hr, color = light_dark))+
  geom_point() +
  facet_wrap(~treatment, scales = "free_x")
```

