---
title: Tile Community Composition Analysis 
author: "Micaela Chapuis"
date: "2026-10-03"
output: github_document
---

## Load Libraries
```{r message = FALSE}
library(tidyverse)
library(here)
library(janitor)
```

## Load Data
```{r message = FALSE}
cover <- read_csv(here("Data", "pilot_random_annotations.csv")) %>% select(-c(Aux4, Aux5, Row, Column)) %>% clean_names() # remove columns I don't need and clean column names
tile_measurements <- read_csv(here("Data", "Respo", "tile_measurements.csv"))
```

Join with tile measurements and filter out non-respo tiles
```{r}
cover <- left_join(cover, tile_measurements, by = c("tile_id" = "tile_ID")) %>% 
  filter(treatment %in% c("white", "black")) %>%
  select(-c("sample_ID", "chamber_volume_ml", "notes"))
```

Edit tile ID to have letter at the beginning so on plots white tiles are always first
```{r}
cover <- cover %>%
  mutate(ID_tile = str_c(str_sub(tile_id, -1), str_sub(tile_id, 1, -2))) # str_c joins two strings, the first one is just the letter (taking the last character out of each tile_id string) and the second one is the numbers (takings everything from the first character to the second to last charafcter)
```


Editing labels
```{r}
cover <- cover %>% mutate(label_code = replace_values(label_code,
                                                      "BTurf" ~ "Brown Turf",
                                                      "BEA" ~ "Brown Encrusting",
                                                      "Unk" ~ "Unknown",
                                                      "ABREV" ~ "Asteronema breviarticulatum", 
                                                      "GEA" ~ "Green Encrusting",
                                                     # "CCA"  not including it retains original value
                                                      "Outside" ~ "Outside Area",
                                                      "UnkMacroa" ~ "Unknown Macroalgae",
                                                      "RedTurf" ~ "Red Turf",
                                                      "Bare-Subst" ~ "Bare Substrate",
                                                      "GTurf" ~ "Green Turf",
                                                      "Unk_Inv" ~ "Unknown Invert"))
```

Calculating percent cover by label
```{r}
percent_cover <- cover %>% 
  filter(!(label_code %in% c("Outside Area"))) %>% # filter out points outside of the settlement area
  group_by(date, site, tile_id, ID_tile, treatment, deployment_block, label_code) %>%
  summarise(num_points = n()) %>% # calculate number of points for each label
  mutate(total_points = sum(num_points), # create a column with the total points on each tile
         percent_cover = num_points/total_points*100) # calculate percent cover for each label
```


```{r fig.width = 10}
percent_cover %>% 
  ggplot(aes(x= ID_tile,
               y = percent_cover, 
               fill= label_code, 
               color= label_code)) + # set lines surrounding each color to match the fill colors 
      geom_bar(stat="identity") + # stacked bars
      facet_wrap(~treatment, scales = "free_x") + 
  
      labs(x = "Tile", # labels
           y="Percent Cover",
           fill = "Category") +

      theme_minimal() +  # theme
      theme(title = element_text(size = 18, face = "bold"), # make title bigger and bold
            axis.title = element_text(size = 16), # make all text bigger
            axis.text = element_text(size = 14),
            axis.text.x = element_text(angle = 45, size = 10),
            legend.title = element_text(size = 16),
            legend.text = element_text(size = 14)) +
    guides(color = "none")
```

```{r fig.width = 10}
percent_cover %>% 
  ggplot(aes(x= ID_tile,
               y = percent_cover, 
               fill= label_code, 
               color= label_code)) + # set lines surrounding each color to match the fill colors 
      geom_bar(stat="identity") + # stacked bars
      facet_grid(~deployment_block, scales = "free_x") + 
  
      labs(x = "Tile", # labels
           y="Percent Cover",
           fill = "Category") +

      theme_minimal() +  # theme
      theme(title = element_text(size = 18, face = "bold"), # make title bigger and bold
            axis.title = element_text(size = 16), # make all text bigger
            axis.text = element_text(size = 14),
            axis.text.x = element_text(angle = 45, size = 10),
            legend.title = element_text(size = 16),
            legend.text = element_text(size = 14)) +
    guides(color = "none")
```


Filtering out just producers
```{r}
producer_cover <- cover %>% 
  filter(!(label_code %in% c("Outside Area", "Bare Substrate", "Unknown"))) %>%
  group_by(date, site, tile_id, ID_tile, treatment, deployment_block, label_code) %>%
  summarise(num_points = n()) %>% # calculate number of points for each label
  mutate(total_points = sum(num_points), # create a column with the total points on each tile
         percent_cover = num_points/total_points*100) # calculate percent cover for each label
```

```{r fig.width = 10}
producer_cover %>% 
  ggplot(aes(x= ID_tile,
               y = percent_cover, 
               fill= label_code, 
               color= label_code)) + # set lines surrounding each color to match the fill colors 
      geom_bar(stat="identity") + # stacked bars
      facet_wrap(~treatment, scales = "free_x") + 
  
      labs(x = "Tile", # labels
           y="Percent Cover",
           fill = "Category") +

      theme_minimal() +  # theme
      theme(title = element_text(size = 18, face = "bold"), # make title bigger and bold
            axis.title = element_text(size = 16), # make all text bigger
            axis.text = element_text(size = 14),
            axis.text.x = element_text(angle = 45),
            legend.title = element_text(size = 16),
            legend.text = element_text(size = 14)) +
    guides(color = "none")
```

## Assign to broader categories
```{r}
producer_cat <- cover %>%
  filter(!(label_code %in% c("Outside Area"))) %>% # filter out points outside of the settlement area
  mutate(category = case_when(
    label_code %in% c("Brown Encrusting", "Green Encrusting") ~ "Encrusting Algae",
    label_code %in% c("Brown Turf", "Red Turf", "Green Turf") ~ "Turf",
    label_code %in% c("Asteronema breviarticulatum", "Unknown Macroalgae") ~ "Macroalgae",
    label_code == "CCA" ~ "CCA",
    label_code %in% c("Bare Substrate", "Unknown") ~ "Other")) %>%
  group_by(date, site, tile_id, ID_tile, treatment, deployment_block,  category) %>%
  summarise(num_points = n()) %>% # calculate number of points for each category
  mutate(total_points = sum(num_points), # create a column with the total points on each tile
         percent_cover = num_points/total_points*100) #%>% # recalculate percent cover for each category, total points still the same as before
  #filter(!category == "Non-Producers")
```

```{r fig.width = 10}
producer_cat %>% 
  ggplot(aes(x= ID_tile,
               y = percent_cover, 
               fill= category, 
               color= category)) + # set lines surrounding each color to match the fill colors 
      geom_bar(stat="identity") + # stacked bars
      facet_wrap(~treatment, scales = "free_x") + 
  
      labs(x = "Tile", # labels
           y="Percent Cover",
           fill = "Category") +

      theme_minimal() +  # theme
      theme(title = element_text(size = 18, face = "bold"), # make title bigger and bold
            axis.title = element_text(size = 16), # make all text bigger
            axis.text = element_text(size = 14),
            axis.text.x = element_text(angle = 45),
            legend.title = element_text(size = 16),
            legend.text = element_text(size = 14)) +
    guides(color = "none") +
  scale_fill_manual(values = c("violetred2", '#FFA07A', "#6E8B3D", "gray45", "#4F94CD")) +
  scale_color_manual(values = c("violetred2", '#FFA07A', "#6E8B3D", "gray45", "#4F94CD"))
```
```{r fig.width = 10}
producer_cat %>% 
  ggplot(aes(x= ID_tile,
               y = percent_cover, 
               fill= category, 
               color= category)) + # set lines surrounding each color to match the fill colors 
      geom_bar(stat="identity") + # stacked bars
      facet_grid(~deployment_block, scales = "free_x") + 
  
      labs(x = "Tile", # labels
           y="Percent Cover",
           fill = "Category") +

      theme_minimal() +  # theme
      theme(title = element_text(size = 18, face = "bold"), # make title bigger and bold
            axis.title = element_text(size = 16), # make all text bigger
            axis.text = element_text(size = 14),
            axis.text.x = element_text(angle = 45),
            legend.title = element_text(size = 16),
            legend.text = element_text(size = 14)) +
    guides(color = "none") +
  scale_fill_manual(values = c("violetred2", '#FFA07A', "#6E8B3D", "gray45", "#4F94CD")) +
  scale_color_manual(values = c("violetred2", '#FFA07A', "#6E8B3D", "gray45", "#4F94CD"))

ggsave(here("Output", "block_percent_cover.jpeg"))

```

Producer categories by treatment
```{r}
producer_cat %>%
  ggplot(aes(x= treatment,
               y = percent_cover, 
               fill= category, 
               color= category)) + # set lines surrounding each color to match the fill colors 
      geom_bar(stat="identity", position = "fill") + 
    
      labs(x = "Treatment", # labels
           y="Percent Cover",
           fill = "Category") +

      theme_minimal() +  # theme
      theme(title = element_text(size = 18, face = "bold"), # make title bigger and bold
            axis.title = element_text(size = 16), # make all text bigger
            axis.text = element_text(size = 14),
            legend.title = element_text(size = 16),
            legend.text = element_text(size = 14)) +
    guides(color = "none") +
  scale_fill_manual(values = c("violetred2", '#FFA07A', "#6E8B3D", "gray45", "#4F94CD")) +
  scale_color_manual(values = c("violetred2", '#FFA07A', "#6E8B3D", "gray45", "#4F94CD"))

ggsave(here("Output", "treatment_percent_cover.jpeg"))
```

## Encrusting vs Turf
Switching to just encrusting/turf (encrusting includes CCA and turf will include macro, which for now is just asteronema and other unidentified small macroalgae)
```{r}
encr_turf <- cover %>%
  filter(!(label_code %in% c("Outside Area"))) %>% # filter out points outside of the settlement area
  mutate(category = case_when(
    label_code %in% c("Brown Encrusting", "Green Encrusting", "CCA") ~ "Encrusting Algae",
    label_code %in% c("Brown Turf", "Red Turf", "Green Turf", "Asteronema breviarticulatum", "Unknown Macroalgae") ~ "Turf",
    label_code %in% c("Bare Substrate", "Unknown") ~ "Other")) %>%
  group_by(date, site, tile_id, treatment, deployment_block, category) %>%
  summarise(num_points = n()) %>% # calculate number of points for each category
  mutate(total_points = sum(num_points), # create a column with the total points on each tile
         percent_cover = num_points/total_points*100) %>% # recalculate percent cover for each category, total points still the same as before
  filter(!category == "Non-Producers")
```

```{r}
ratio <- encr_turf %>%
  select(-c("num_points", "total_points")) %>%
  pivot_wider(
    names_from = "category",
    values_from = "percent_cover") %>%
  mutate(ratio = `Encrusting Algae`/Turf)
```

```{r}
ratio %>% write.csv(here("Data", "encrusting_turf_ratio.csv"))
```


### Tests

#### Ratio black vs white
```{r} 
ratio %>% ggplot(aes(x = treatment, y = ratio)) + geom_boxplot()
```

```{r}
# reshape data so each block has a white ratio and a black ratio
block_ratio <- ratio %>% ungroup() %>% select(deployment_block, treatment, ratio) %>%
  pivot_wider(names_from = treatment, values_from = ratio)

block_ratio
```

```{r}
# test whether black tiles have higher encr/turf ratio than white tiles 
# NOT paired (treats black and white tiles as independent samples, is the avg ratio different between all black and all white tiles?)
t.test(block_ratio$black, block_ratio$white)
```

```{r}
# test whether black tiles have higher encr/turf ratio than white tiles 
# paired t-test (is the ratio higher on black tiles than on their paired white tile)
t.test(block_ratio$black, block_ratio$white, paired = TRUE)
```


#### Encrusting black vs white
```{r} 
ratio %>% ggplot(aes(x = treatment, y = `Encrusting Algae`)) + geom_boxplot()
```
```{r}
# reshape data so each block has a white encrusting and a black encrusting
block_encr <- ratio %>% ungroup() %>% select(deployment_block, treatment, `Encrusting Algae`) %>%
  pivot_wider(names_from = treatment, values_from = `Encrusting Algae`)

block_encr
```


```{r}
# test whether black tiles have higher encrusting cover than white tiles 
# NOT paired (treats black and white tiles as independent samples, is the avg encr cover different between all black and all white tiles?)
t.test(block_encr$black, block_encr$white)
```

```{r}
# test whether black tiles have higher encrusting cover than white tiles 
# paired t-test (is the encrusting cover higher on black tiles than on their paired white tile)
t.test(block_encr$black, block_encr$white, paired = TRUE)
```

#### Turf black vs white
```{r}
ratio %>% ggplot(aes(x = treatment, y = Turf)) + geom_boxplot()
```

```{r}
# reshape data so each block has a white turf and a black turf
block_turf <- ratio %>% ungroup() %>% select(deployment_block, treatment, Turf) %>%
  pivot_wider(names_from = treatment, values_from = Turf)

block_turf
```


```{r}
# test whether black tiles have higher turf cover than white tiles 
# NOT paired (treats black and white tiles as independent samples, is the avg turf cover different between all black and all white tiles?)
t.test(block_turf$black, block_turf$white)
```

```{r}
# test whether black tiles have higher turf cover than white tiles 
# paired t-test (is the turf cover higher on black tiles than on their paired white tile)
t.test(block_turf$black, block_turf$white, paired = TRUE)
```